---
title: "Preprocessing"
author: "Mikhail Dozmorov"
output:
  html_document:
    toc: true
    theme: united
date: "`r Sys.Date()`"
---

Material is public domain

```{r setup, echo=FALSE}
# Set up the environment
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=T, tidy=T, fig.keep='high', echo=T, dpi=100, out.width=700)
options(replace.assign=TRUE, width=220)
options(stringsAsFactors = FALSE)
set.seed(1)
```

```{r echo=FALSE}
# http://jura.wi.mit.edu/bio/education/bioinfo2007/arrays/array_exercises_1R.html
# http://bioconductor.org/packages/release/bioc/vignettes/annotate/inst/doc/annotate.pdf
# http://manuals.bioinformatics.ucr.edu/home/R_BioCondManual#biocon_affypack
# http://www2.warwick.ac.uk/fac/sci/moac/people/students/peter_cock/r/geo/
```

# Microarray data retrieval and preprocessing

## Introduction

You'll be using a sample of expression data from a study using Affymetrix (one color) U95A arrays that were hybridized to tissues from fetal and human liver and brain tissue. Each hybridization was performed in duplicate. Many other tissues were also profiled but won't be used for these exercises.

### Preliminary steps: Image analysis and calculation of expression value

As described in [Su et al., 2002](https://www.ncbi.nlm.nih.gov/pubmed/11904358?dopt=Abstract), human tissue samples were hybridized on Affymetrix (one-color) arrays and chips were scanned. For each tissue, at least two independent samples were hybridized to separate chips.

Scanned images were quantified (including measurement of background) using standard software.

### Preprocessing and normalization of Affymetrix expression data

Data from a probeset (a series of oligos designed to a specific gene target) needs to be summarized by calculating an expression values for that probeset. This can be done using a Bioconductor/R version of the methods in the Microarray Suite 5.0 (MAS5, the standard Affymetrix algorithm) used by the Affymetrix software suite.

See the [manuals](http://www.affymetrix.com/support/technical/manuals.affx) from Affymetrix for more information about these processes, and the [Statistical Algorithms Description Document](http://www.affymetrix.com/support/technical/whitepapers/sadd_whitepaper.pdf) for the actual equations used.

You also have the option of performing other algorithms such as RMA, GCRMA, and dChip.

[Download](http://jura.wi.mit.edu/bio/education/bioinfo2007/arrays/Su_CELs.zip) the starting data ("CEL" files) and unzip this file to get 8 files ending in ".CEL".

```{r eval=FALSE}
# Install Bioconductor, if necessary
source("http://bioconductor.org/biocLite.R") # Import biocLite() function into R environment
biocLite() # Install or update Bioconductor
```

Install, if necessary, and load the `affy` library that contains the Affymetrix microarray functions
 
```{r}
# biocLite("affy") # Install affy package
library(affy) # Load affy package
```

Read the CEL files (first command below), save expression values as a tab-separated text file, and then summarize and normalize with MAS5 (second command below). This could take a few minutes.

```{r}
affy.data = ReadAffy(celfile.path="../data/Su_CELs/")
eset.mas5 = mas5(affy.data) # Summarize with MAS5 algorithm
```

The variable 'eset.mas5' contains normalized expression values for all probesets, along with other information. Let's check its parameters

```{r}
eset.mas5 # A summary of AffyBatch object
head(exprs(eset.mas5)) # Peek on the expression matrix
pData(eset.mas5) # Check experimental phenoData
```

The phenodata object contains all the information about the samples of our dataset. It is always important to keep track off all the information about our samples, since they can strongly affect the final results.

It is possible to visually explore the dataset using different methods.
The boxplot visualizes the distribution of the intensities in each array of the dataset.

```{r}
boxplot(log2(exprs(eset.mas5))) # Boxplot of log2-transformed data
```

It is also possible to visualize the image representing each CEL file (in this example, we visualize the first slide of the dataset only). In this way, it is possible to pinpoint technical problems occurring eventually only to one region of the array.

```{r}
image(affy.data[1]) # Check image of the first CEL file
```

Let's continue by getting the expression matrix (probesets/genes in rows, chips in columns).

```{r}
exprSet.nologs = exprs(eset.mas5) # Expression matrix
colnames(exprSet.nologs) # List the column (chip) names
# Rename the column names if we want
colnames(exprSet.nologs) = c("brain.1", "brain.2", 
                             "fetal.brain.1", "fetal.brain.2",
                             "fetal.liver.1", "fetal.liver.2", 
                             "liver.1", "liver.2")
```

At this time let's log-transform the expression values to get a more normal distribution. We have to remember we've done this when we calculate ratios. Logarithms can use any base, but base 2 is easiest when transforming ratios, since transformed 2-fold ratios up or down will be +1 or -1. As a result, we'll do all logs with base 2 to keep thing simplest.

```{r}
exprSet = log(exprSet.nologs, 2) # Log2-transformed expression
```

Optional. In place of mas5 normalization above, we could choose many other ways to preprocess the data, such as RMA (which outputs log2-transformed expression values)

```{r eval=FALSE}
eset.rma <- justRMA(celfile.path="../data/Su_CELs/") # RMA summarization of the CEL files
```

or GCRMA (which also outputs log2-transformed expression values)

```{r eval=FALSE}
biocLite("gcrma") # Install GCRMA package, if absent
library(gcrma) # Load GCRMA package
eset.rma <- justGCRMA(celfile.path="../data/Su_CELs/") # GCRMA summarization of the CEL files
```

or dChip (also known as MBEI; not log-transformed)

```{r eval=FALSE}
eset.dChip = expresso(affy.data, normalize.method="invariantset", 
                      bg.correct=FALSE, pmcorrect.method="pmonly",summary.method="liwong" # dCHIP summarization
```
					  
To print out our expression matrix (as with most data), we can use a command like

```{r eval=FALSE}
write.table(exprSet, file="Su_mas5_matrix.txt", quote=F, sep="\t")
```

to get a tab-delimited file that we could view in Excel or a text editor.
While we're doing Affymetrix-specific preprocessing, let's calculate an Absent/Present call for each probeset.
 
Run the Affy A/P call algorithm on the CEL files we processed above

```{r}
data.mas5calls = mas5calls(affy.data)
# Get the actual A/P calls
data.mas5calls.calls = exprs(data.mas5calls)
# What's a typical A/M/P distribution?
table(data.mas5calls.calls[, 1])
# Print the calls as a matrix
# write.table(data.mas5calls.calls, file="Su_mas5calls.txt", quote=F, sep="\t")
```

Other methods of normalization available: ?justPlier, ?expresso

### Normalization of expression data

Why normalize? Chips may have been hybridized to different amounts of RNA, for different amounts of time, with different batches of solutions, etc. Normalization should remove systematic biases and make any comparisons between chips more meaningful.
If you performed Part 0 above, your data is already normalized. But if you got an unnormalized expression from another source, how could you normalize it?

```{r}
# exprSetRaw <- read.delim("http://jura.wi.mit.edu/bio/education/bioinfo2007/arrays/Su_raw_matrix.txt", sep = "\t")
exprSetRaw = read.delim("../data/Su_raw_matrix.txt", sep="\t") # Set separator as tab
```

Calculate the trimmed mean of all expression values on each chip.

A trimmed mean calculates a summary value that is somewhere between the mean and median of the set of values.

We remove the top and bottom 2% of values (for example), and find the mean of the remaining values. To get the trimmed mean of column 1 on our matrix, we can use a command like

```{r}
trmean.col.1 = mean(exprSetRaw[,1], trim=0.02) # Trimmed mean in the first column
```

We can apply this command to all columns at once with the "apply" command, where the '2' means to run the command on all columns (and '1' would be to do the same thing on all rows).

```{r}
trmean = apply(exprSetRaw, 2, mean, trim=0.02) # Trimmed mean in all columns
trmean
```

If we're curious about the variation between genes on each chip, we can find the standard deviation for each chip

```{r}
sd = apply(exprSetRaw, 2, sd)
sd
```

Or we could compare medians

```{r}
median = apply(exprSetRaw, 2, median)
median
```

Divide all expression values by the mean for that chip, and multiply by a scaling factor, such as the mean of the trimmed means.

```{r}
mean.of.trmeans = mean(trmean)
exprSet.trmean = exprSetRaw / trmean * mean.of.trmeans
```

Let's look at the data before and after our normalization

```{r}
op <- par()
par(mfrow = c(1, 2))
boxplot(log2(exprSetRaw)) # Data before
boxplot(log2(exprSet.trmean)) # And after trimmed mean normalization
par(op)
```

Print the data that's been normalized by the global trimmed mean

```{r eval=FALSE}
write.table(exprSet.trmean, file="Su_mas5_trmean_norm.txt", quote=F, sep="\t")
```

For further processing, rename your data

```{r}
exprSet = exprSet.trmean
```

Quantile normalization is often too extreme, but it's common for Affymetrix probe-level normalization (being part of MAS5). If you wish to use it, one way is to use a command from the "limma" package

```{r}
# biocLite("limma")
library(limma)
exprSet.quantile = normalizeQuantiles(exprSet)
```

And visualize our work

```{r}
boxplot(log2(exprSet.quantile))
```

Another common normalization choice (generally for 2-color arrays) is loess. We combined the "brain.1" and "fetal.brain.1" expression data into a file that we can pretend is 2-color data (together with 0 values for "background"). Then we can normalize it with loess.

Start by downloading [brain.fetalbrain.2color.data.txt](http://jura.wi.mit.edu/bio/education/bioinfo2007/arrays/brain.fetalbrain.2color.data.txt) into your working directory. Read this fake 2-color file, indicating the columns for the Red and Green channels and the Red and Green backgrounds.

```{r}
brain.fetalbrain.2color = read.maimages("../data/brain.fetalbrain.2color.data.txt", 
	columns=list(G="brain.1", R="fetal.brain.1", Gb="bg1", Rb="bg2"))
```

Now we have the two channels of data which we can loess normalize

```{r}
brain.fetalbrain.2color.loess = 
	normalizeWithinArrays(brain.fetalbrain.2color, method="loess")
```

How do you know if anything changed? Print an MA plot before and after normalization

```{r}
op <- par()
par(mfrow=c(1,2)) # Set up a page with two figures next to each other
limma::plotMA(brain.fetalbrain.2color) # Print the figures
limma::plotMA(brain.fetalbrain.2color.loess) # and after loess
par(op)
```

How do the figures differ?

